---
title: "Developing Data Products Week 4 Project 1"
author: "jedraynes"
date: "4/2/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
pacman::p_load(tidyverse, 
               plotly, 
               ggplot2)
```

## Project Overview

For this project, I created an RShiny application that plots the prices of Bitcoin (BTC), Ethereum (ETH), and the S&P 500 index (using the ETF SPY as a proxy). The data is sourced from Yahoo Finance.

## Interactivity

To satisfy the interactivity requirement, I added three main features:

- Ticker selection: The user can select one or multiple tickers to plot on the graph.
- Date input: The maximum range is the last 5 years since the data was downloaded on April 2, 2021. You can set your own date range to see what the price was historically during that range.
- Plotly: The chart itself is interactive as it was built using 'ggplot2' and then presented using 'plotly'. Users can hover over the line to see the actual value.

The next slide shows the output for all dates and all tickers.

## Slide with Plot

```{r, echo = FALSE}
# setwd
setwd('C:\\Users\\Jed\\iCloudDrive\\Documents\\Learn\\R\\Johns Hopkins Data Science Specialization\\9 Developing Data Products\\Week 4\\Developing_Data_Products_Week_4_Project_1\\data')

# load data
project1_data <- read.csv('project1_data.csv')
project1_data$Date <- as.Date(project1_data$Date, format = '%Y-%m-%d')

# plot
ggplotly(ggplot(project1_data) +
      geom_line(aes(x = Date, y = value, color = variable)) +
      xlab('Date') +
      ylab('Value ($)') +
      theme_bw() +
      theme(panel.background = element_blank(),
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(), 
            legend.position = 'bottom', 
            legend.title = element_blank(), 
            plot.title = element_text(face = 'bold', hjust = 0.5)))
```

## Conclusion

Thank you!
